\documentclass[11pt]{article}

\usepackage[nohide,twocolumn]{ulecnot}
\usepackage{natbib}


\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}

\usepackage{hyperref}


\pagestyle{fancy}
\lhead{COGS 501 -- Formal Languages and Linguistics}
\chead{Formal Language Basics and Regular Languages}
	\rhead{Updated \it \today }
\lfoot{Umut \"Ozge}
\cfoot{}
\rfoot{Page \thepage/\pageref{LastPage}}
\setlength{\headheight}{13.6pt}

\newcommand{\emptystring}{\ensuremath{\epsilon}}


\begin{document}
\tableofcontents

\section{Alphabets,  strings, languages}

\ezimeti{

\item An \uterm{alphabet} is a finite set of \uterm{symbols}.

\item[] E.g.\ the Roman alphabet $\{a,\ldots,z\}$, $\{a,b\}$, $\crbr{0,1}$,\ldots

\item A \uterm{string} over an alphabet is a finite sequence of symbols from that alphabet.

\item The \uterm{empty string} consists of zero symbols. We will denote it by the symbol
`\emptystring'.
% %\footnote{Sudkamp uses `$\lambda$'  for the empty string, you may 
% also encounter `$e$' or other symbols in other books. As `$\lambda$' has another
% ubiquitous use in computer science and linguistics, I will use `\emptystring'.}

\item[] Examples of strings on alphabet $\{a,b\}$ are \emptystring, $a$,
$abbaba$, \ldots
\item[] The symbols $u,v,w,x,y,z$ are used to name strings -- therefore we avoid
them as symbols of alphabets.

\item The set of all strings -- including \emptystring -- over an alphabet $\Sigma$ is denoted by \sigmastar.

\item The \uterm{length} of a string $w$ is denoted by $|w|$.

\item The \uterm{concatenation} of two strings $w$ and $v$ is formed by
sequencing the strings in the given order; it is denoted as $wv$, $w\circ v$,
or $w^\frown v$. Concatenation is associative: $(xy)z = x(yz)$, and
$\emptystring w = w \emptystring = w$.


\item A string $v$ is a \uterm{substring} of a string $w$, if there exists
strings $x$ and $y$ such that $w=xvy$. Either or both of $x$ and $y$ can be
\emptystring. 

\item The \uterm{reverse} of a string $w$, denoted as $w^R$, is defined as
follows:

\begin{udefinition}[Reverse of a string]
\ezimeti{
\item[]
\item[i.] If $w = \emptystring$, then $w^R = w$.
\item[ii.] If $w = va$ for some $a \in \Sigma$, then $w^R = av^R$.
}
\end{udefinition}

\item The notation $w^n$ stands for concatenating $w$ to itself for $n$ times.
$w^0 = \emptystring$, $w^1 = w$, $w^2 = ww$, and so on.

\item A \uterm{language} is a set of strings over a certain alphabet.

\item Therefore a language $L$ on an alphabet $\Sigma$ is a subset of
\sigmastar.

\item Some example languages over $\Sigma = \{a,b\}$:

\[
\{b,aa,ab\}
\]
\[
\setabs{w\in \sigmastar}{w \text{ has equal number of }a\text{'s and
}b\text{'s}}\\
\]
\[
\setabs{w\in \sigmastar}{w=w^R}\\
\]

\hrulefill
\begin{uexercise}\label{whichislang}
Which of the following are languages?
\[
\emptystring\quad\quad\{\emptystring\}\quad\quad\emptyset\quad\quad\Sigma\quad\quad\sigmastar
\]

\hyperlink{whichislangsol}{\qed}
\end{uexercise}
\hrulefill



}



\section{Some operations on languages}

\ezimeti{ 

\item Given that languages are sets, ordinary set operations
\cttr{union}, \cttr{intersection} and \cttr{difference} are defined
for languages. 

\item There also are operations specific to languages. One is
\cttr{concatenation of languages}. Given any languages $L_1$ and $L_2$
over $\Sigma$, their concatenation, designated as 
$L_1\circ L_2$, $L_1^\frown L_2$, or simply $L_1L_2$, is defined as
follows:
\begin{align}
L_1L_2=\{
w \in \Sigma^*\ |\ 
w = xy\text{, for some } x\in L_1 \text{ and } y\in L_2
\}
\end{align}

\item Our final and third operation is \textbf{closure} (or
\textbf{star}, or \textbf{Kleene closure}) of a language $L$, denoted
as $L^*$, which is the set of expressions formed by concatenating zero or more
strings from $L$.  Formally, 
\[
L^*=\setabs{w\in \sigmastar}{w=w_1w_2\ldots w_k \text{ for some }k\geq 0 \text{ where }w_1,w_2,\ldots w_k \in L}
\]
or,
\begin{align*}
&L^* = \bigcup_{i = 0}^\infty L^i \\ & \text{where }
L^0=\crbr{\epsilon},\ L^1=L,\ \text{ and, } L^i=LL\cdots L\text{,
with } i\text{-many } L\text{s} &
\end{align*}


\item We write $L^+$ in place of $LL^*$, which is:
\[
L^+=\setabs{w\in \sigmastar}{w=w_1w_2\ldots w_k \text{ for some }k\geq 1 \text{ where }w_1,w_2,\ldots w_k \in L}
\]


\hrulefill
\begin{uexercise}\label{extf1}
State whether true or false:
\begin{enumerate}
\item\label{extf1a} $\{\epsilon\}^* = \{\epsilon\}$
\item\label{extf1b} $\emptyset^*=\{\epsilon\}$
\item\label{extf1c} For any alphabet $\Sigma$, any $L$ defined over $\Sigma$ is such that
$L\in \mathcal{P}(\Sigma^*)$. ($\mathcal{P}(X)$ denotes the power set of $X$.)
\item\label{extf1cc}  For any language $L$, $L^* = (L^*)^*$.
\item\label{extf1d} For any language $L$, $\emptyset L = L\emptyset = L$
\item\label{extf1e} For any language $L$, $\{\epsilon\}L = \emptyset$
\end{enumerate}

\hyperlink{extf1sol}{\qed}
\end{uexercise}

% \begin{uexercise}\label{write-ext} Give $L^*$ for (i) $L=\{0,1\}$;  (ii) $L$ is the set of
% 	strings of 0's, and (iii) $L=\emptyset$. 
% \end{uexercise}

\begin{uexercise}\label{exconcuni}
Let $L_1 = \setabs{w \in \{ a,b\}^*}{|w| = 2}$ and $L_2\setabs{w \in \{ a,b\}^*}{|w| = 3 \text{ and } w \text{ ends with } b}$

\begin{enumerate}
\item\label{exconcunia} Give the concatenation of $L_1$ and $L_2$.
\item\label{exconcunib} Give their union.
\end{enumerate}

\hyperlink{exconcunisol}{\qed}
\end{uexercise}



\begin{uexercise}\label{lang-desc}
Let $L_1=\setabs{w \in \crbr{\text{0,1}}^*}{w \text{ has an even number of
0's}}$ and  \\
$L_2= \setabs{w \in \crbr{\text{0,1}}^*}{w \text{ starts
with a 0 followed by any number of 1's}}$. \\ Which language is $L_1L_2$?

\hyperlink{lang-desc-sol}{\qed}
\end{uexercise}
\hrulefill

}

\section{Finite representation of languages}

\ezimeti{
\item In the theory of computation and its applications we are interested in
representing languages of our interest with \emph{finite} means. This is easy
when the language is finite, but it is a challenge for nonfinite
languages.

\item One method is constructing an \uterm{inductive} definition:


\begin{uexample}[Inductive definition of a language]
The language $L$ over $\{a,b\}$, where each string begins with an $a$ and has
an even length.
\ezimeti{
\item[i.] $aa$ and $ab \in L$.
\item[ii.] If $w \in L$, then $waa$, $wab$, $wba$, $wbb \in L$.
\item[iii.] Nothing other than the strings obtained via i.\ and ii.\ above are
in $L$.
}
\end{uexample}

\hrulefill
\begin{uexercise}\label{induct}
Write an inductive definition for the language $L$ over $\crbr{a,b}$ in which
every occurrence of $b$ is immediately preceded by an $a$.

\hyperlink{induct-sol}{\qed}
\end{uexercise}
\hrulefill

\item Now let us see a more transparent and direct way of specifying
the above languages. This method involves applying the operations set union, concatenation and
closure on sets. 

\begin{uexample}
The language $L$ over $\{a,b\}$ which has $bb$ as a substring can be defined as
$\{a,b\}^*\{bb\}\{a,b\}^*$.
\end{uexample}



\hrulefill
\begin{uexercise}\label{set-notation}
\begin{enumerate}
\item[]
\item\label{set-not1} Define the language $L$ over $\{a,b\}$ whose strings either start with $aa$ or end with $bb$.
\item Define the language $L$ over $\{a,b\}$ whose strings have an even
length. Also define for odd length.
\item Define the language $L$ over $\{0,1\}$ whose strings have two or three
occurrences of 1 the second and third of which are not consecutive.
\end{enumerate}

\hyperlink{set-notationsol}{\qed}
\end{uexercise}
\hrulefill


}

\section{Regular languages}

\ezimeti{

\item Another central point of interest in the theory of computation is classes
of languages -- the set of all languages that share a certain mathematically specifiable
property.


\item The first class we will look at is the class (or set) of \uterm{regular languages}
(or \uterm{regular sets}).



\begin{udefinition}[Regular Languages]\label{def-reglang}
Given an alphabet $\Sigma$:

\etaremune{
\item $\emptyset$ is a regular language.
\item For any symbol $a \in \Sigma$, $\crbr{a}$ is a regular language.
\item If $A$ and $B$ are regular languages, so is $A\cup B$.
\item If $A$ and $B$ are regular languages, so is $AB$. 
\item If $A$ is a regular language, so is $A^*$.
\item Nothing is a regular language unless it fits the above
definition.
}
\end{udefinition}

\item In other words, a language is regular if it can be constructed
from unit languages like $\{a\}$, $\{b\}$ etc.\  and the empty
language $\emptyset$ by the repeated application of
union, concatenation and closure.

\item Precedence conventions: Kleene star binds most tightly, then comes
concatenation, and finally union. For instance, $\crbr{a}\crbr{b}^*$ gives
$\crbr{a,ab,abb,abbb,\ldots}$; if you want
$\crbr{\epsilon,ab,abab,ababab,\ldots}$, you need to have 
$(\crbr{a}\crbr{b})^*$, Again, $\crbr{a}\crbr{b}\cup\crbr{c}$ gives the set
$\crbr{ab,c}$; if you want to have $\crbr{ab,ac}$ you need
$\crbr{a}(\crbr{b}\cup\crbr{c})$. 


\begin{uexercise}\label{show-reg}
Show that the following languages are regular. 
\etaremune{
\item
$L=\setabs{x \in \crbr{a,b}^*}{x \text{ contains an odd number of $b$'s}}$  
\item
$L=\setabs{x \in \crbr{a,b}^*}{x \text{ contains exactly two or three 
$b$'s}}$
}
\end{uexercise}

\item \textbf{Regular expressions} are notational devices to represent
regular languages.

\begin{udefinition}[Regular Expressions]  For each regular expression
$E$,  the language denoted by it is designated as $L(E)$. The set of
regular expressions can be inductively defined as follows.

\etaremune{
\item The constants $\mathbf{\epsilon}$ and $\mathbf{\emptyset}$ are regular
expressions, where $L(\mathbf{\epsilon})=\{\epsilon\}$ and
$L(\mathbf{\emptyset})=\emptyset$.
\item If $a$ is a symbol, $\mathbf{a}$ is a regular expression, where
$L(\mathbf{a})=\{a\}$.
\item If $E$ and $F$ are regular expressions, so is $E\cup F$, where
$L(E\cup F)=L(E)\cup L(F)$.
\item If $E$ and $F$ are regular expressions, so is $EF$, where
$L(EF)=L(E)L(F)$.
\item If $E$ is a regular expression, so is $E^*$, where
$L(E^*)=L(E)^*$
\item If $E$ is a regular expression, so is $(E)$, where $L((E))=L(E)$
\item If $E$ is a regular expression, then it can be shown to be so by 1--6.
}
\end{udefinition}


\item Same precedence conventions as above apply.

\begin{uexercise}\label{writere} Write regular expressions for the following languages:
\begin{enumerate}
\item\label{writerea} The set of strings over alphabet $\{a,b,c\}$ containing at least one $a$ and at least one $b$.
\item\label{writereb} The set of strings that consist of alternating 0's and 1's (=strings with
no consecutive 0's or 1's).
\item\label{writerec} The set of strings of 0's and 1's whose third symbol from the right end is 1.
\item\label{writered} The set of strings of 0's and 1's with at most one pair of consecutive 1's.
\item\label{writeree} The set of strings of 0's and 1's with no substring 111. 
\item\label{writeref} $\setabs{w\in \{a,b\}^*}{w \text{ has at least three } a\text{'s} }$.
\item\label{writereg} $\setabs{w\in \{a,b\}^*}{w \text{ has at least three } a\text{'s or
 at least
two } b\text{'s}}$.
\item\label{writereh} $\{w\in \{0,1\}^*\, |\, w \text{ starts with 0 and has
				odd length, or starts with 1}$ and has even length$\}$.

\item\label{writerei} $\setabs{w\in \{0,1\}^*}{w\text{ starts and ends with the same symbol}}$.

\item\label{writerej} $\setabs{w\in \{a,b\}^*}{w \text{ does not begin with } aaa}$.

\item\label{writerek} $\setabs{w\in \{a,b,c\}^*}{w \text{ has an odd number of occurrences of } ab}$.

\item\label{writerel} $\setabs{w\in \{a,b\}^*}{w \text{ has an even length and odd number of
}b\text{'s}}$.
\end{enumerate}
	
\hyperlink{writeresol}{\qed}
\end{uexercise}

}

\appendix
		\section{Answers for selected exercises}
\begin{itemize}

\item[\ref{whichislang}]\hypertarget{whichislangsol}{} 

$\emptystring$ is a string, not a language; $\{\emptystring\}$,
$\emptyset$, $\Sigma$, $\sigmastar$ are all
sets of strings, therefore are all languages. 

\item[\ref{extf1}]\hypertarget{extf1sol}{} \ref{extf1a}: T, \ref{extf1b}: T, \ref{extf1c}: T, \ref{extf1cc}: T,\ref{extf1d}: F, \ref{extf1e}: F

\item[\ref{exconcuni}]\hypertarget{exconcunisol}{}

\ref{exconcunia}.
$L_1L_2= \{aaaab,aaabb,aabab,aabbb,abaab,ababb,abbab,abbbb,$ 

$baaab, baabb, babab, babbb, bbaab,bbabb,bbbab,bbbbb\}$

\ref{exconcunib}. $L_1\cup L_2 =\crbr{
aa,
ab,
bb,
ba,
aab,
abb,
bbb,
bab}
$

% \item[\ref{write-ext}] (i) $L^*= \crbr{0,1}^*$; (ii) $L^*=\crbr{0}^*$; (iii)
% $L^*=\crbr{\epsilon}$.

\item[\ref{lang-desc}]\hypertarget{lang-desc-sol}{} Strings with odd number of 0's. 

\item[\ref{induct}]\hypertarget{induct-sol}{}

\ezimeti{
\item[i.] $\epsilon \in L$.
\item[ii.] If $w \in L$, then $wab$ and $wa \in L$.
\item[iii.] Nothing other than the strings obtained via i.\ and ii.\ above are
in $L$.
}

\item[\ref{set-notation}]\hypertarget{set-notationsol}{}
\begin{enumerate}
\item  $(\{aa\}\{a,b\}^*)\cup(\{a,b\}^*\{bb\})$;

\item even length:
$\{ab,bb,ba,ab\}^*$ or $(\{a,b\}\{a,b\})^*$;

odd length:
$\{ab,bb,ba,ab\}^*\{a,b\}$ or $(\{a,b\}\{a,b\})^*\{a,b\}$;

\item $\crbr{0}^*\crbr{1}\crbr{0}^*\crbr{1}\crbr{0}^*\crbr{01,\epsilon}\crbr{0}^*$.

\end{enumerate}

\item[\ref{show-reg}] We need to show that the given languages can be defined
according to Definition~\ref{def-reglang}. There are two tricky points. One,
you can only use unit languages in your solution, you are not allowed to
use sets like $\crbr{a,b}$, you need to obtain them by union. Two,  you are not
given the empty string $\epsilon$ in the definition, so you cannot use the unit
language $\crbr{\epsilon}$  in your solution. However, you can obtain it from
the empty set, since $\emptyset^*=\crbr{\epsilon}$.

\begin{enumerate}
\item $\crbr{a}^*\crbr{b}\crbr{a}^*(\crbr{a}^*\crbr{b}\crbr{a}^*\crbr{b}\crbr{a}^*)^*$ 

\item $\crbr{a}^*\crbr{b}\crbr{a}^*\crbr{b}\crbr{a}^*(\crbr{b}\cup\emptyset^*)\crbr{a}^*$ 

\end{enumerate}


\item[\ref{writere}] \hypertarget{writeresol}{}
\begin{enumerate}
\item[\ref{writerea}.] $((a\cup b \cup c)^*a(a\cup b \cup c)^*b(a\cup b \cup c)^*) \cup ((a\cup b \cup c)^*b(a\cup b \cup c)^*a(a\cup b \cup c)^*)$

or $(a\cup b \cup c)^*((a(a\cup b \cup c)^*b) \cup (b(a\cup b \cup c)^*a))(a\cup b \cup c)^*$

or $(a\cup b \cup c)^*((ac^*b) \cup (bc^*a))(a\cup b \cup c)^*$
\item[\ref{writereb}.] $(1\cup\epsilon)(01)^*(0\cup\epsilon)$

\item[\ref{writerec}.] $(0\cup 1)^*1(0\cup 1)(0\cup 1)$

\item[\ref{writered}.] $(0\cup 10)^*(11\cup 1\cup \epsilon)(0\cup 01)^*$

or $(1\cup \epsilon)(0\cup 01)^*(10 \cup 0)^*(1\cup\epsilon)$

\item[\ref{writeree}.] $(0\cup 10 \cup 110)^*(11\cup 1\cup \epsilon)$

\item[\ref{writeref}.] $(a\cup b)^*a(a\cup b)^*a(a\cup b)^*a(a\cup b)^*$

\item[\ref{writereg}.] $((a\cup b)^*a(a\cup b)^*a(a\cup b)^*a(a\cup b)^*) \cup((a\cup b)^*b(a\cup b)^*b(a\cup b)^*)$

\item[\ref{writereh}.] $(0(10\cup 01\cup 11\cup 00)^*) \cup ((10\cup 11)(10\cup 01\cup 11\cup 00)^*)$
or $(0((1\cup 0)(1\cup 0))^*) \cup ((10\cup 11)((1\cup 0)(1\cup 0))^*)$

\item[\ref{writerei}.] $0(1\cup 0)^*0\cup 1(1\cup 0)^*1 \cup 0 \cup 1$
\item[\ref{writerej}.]
$\epsilon\cup a \cup aa\cup (\epsilon\cup a\cup aa)b(a\cup b)^*$
or $(\epsilon\cup a\cup aa)(b(a\cup b)^* \cup \epsilon)$

\item[\ref{writerek}.] $(a^*c\cup b\cup c)^*a^*ab((a^*c\cup b\cup c)^*a^*ab(a^*c\cup b\cup c)^*a^*ab)^*(a^*c\cup b\cup c)^*a^*$

\item[\ref{writerel}.] 
This question is harder than the others. It's quite normal if you were not
able to solve it. I included it in the exercises, because it exemplifies the
kind of abstract thinking one needs to adopt in dealing with the mathematics of
strings and mathematics in general. In short, the difficulty level of this
question is above what you will encounter in exams, however studying its
solution would be very helpful for understanding and solving simpler questions.

First, let us observe that, as our grammatical strings will be of even length,
they are divisible into 2 symbol chunks -- you can scan each string from left to
right looking at two symbols at each step:

\begin{align}
x_1y_1\, x_2y_2\,\ldots\, x_ny_n
\end{align}

for some $n \geq 1$. The reason we say $n$ needs to be at least 1 is that we
need to have an odd number of $b$'s in our strings and therefore we cannot have
the empty string: there needs to be at least one pair of symbols.

Now let us observe that in every grammatical string there \emph{must} be at
least one pair which is either $ab$ or $ba$. The reason is that, if all were
$aa$ or $bb$ we would have an even number of $b$'s, therefore scanning the pair
of symbols from left to right, we are certain that we will encounter an $ab$ or
$ba$ before the string is over. There may be more than one $ab$ or $ba$, the
crucial thing is that there must be at least one.

As we are sure that there will be at least one $ab$ or $ba$ somewhere in our
grammatical strings, any such string will obey the following incomplete regular
expression:

\begin{align}\label{middle}
\ldots (ab \cup ba)\ldots
\end{align}



Now the question is what should come in places of `$\ldots$'. Think for some
time before reading on, considering that we want our strings to have an even
length and odd number of $b$'s.

As we already have a string of even length and odd $b$'s (namely $ab$ or $ba$),
whatever we put to the right or left of it should not change this property.
Therefore the incomplete parts should contribute strings of even length and an
\emph{even} number of $b$'s, \emph{including the empty string}. If both
substrings on the left and right of $ab$ or $ba$ we already specified in
(\ref{middle}) have this
property, then we will end up with a string of even length and odd number of
$b$'s. Therefore, our task is reduced to writing a regular expression for strings
with even length and even number of $b$'s. Now try to write this on your own and
check with the solution given below.

How can we write a regular expression for strings with an even length and even
number of $b$'s? First observe that, as we don't know the length of those
strings in advance, there must be a repetition with Kleene star. The second
observation is that this repetition should contribute an even number of symbols
including an even number of $b$'s. The strings $aa$ and $bb$ fulfill
this condition, we can repeat them as many times as we want. So we will have
something like $(\epsilon \cup aa \cup bb)^*$ somewhere in our expression. Is
this enough? No, because we may encounter $ab$'s or $ba$'s in our even length
string with even number of $b$'s. We cannot repeat these, since each repetition
contributes and odd number of $b$'s -- namely 1 -- thereby running the risk of
ending up with odd number of $b$'s. Once again, the critical observation is that
when we encounter an $ab$ or $ba$ in the string, we need to have another $ab$ or
$ba$ somewhere down the string, so that we keep the number of $b$'s balanced at
an even number. In oder words when you see an $ab$ or $ba$ you start to owe
another $b$ to be re-paid by another $ab$  or $ba$. This point of repayment does
not need to come immediately after you start owing the $b$, because there may
come $aa$'s and $bb$'s in between. Bringing all these observations together we
arrive at the following regular expression that defines all the strings over the
alphabet $\crbr{a,b}$ with an even length and an even number of $b$'s:

\begin{align}
(aa \cup bb \cup ((ab \cup ba)(aa\cup bb)^* (ab\cup ba)))^*
\end{align}


Bringing all together, we arrive at the desired solution:

\begin{align}
(aa \cup bb \cup ((ab \cup ba)(aa\cup bb)^* (ab\cup ba)))^*(ab \cup ba)(aa \cup
bb \cup \\ ((ab \cup ba)(aa\cup bb)^* (ab\cup ba)))^* \nonumber
\end{align}


Now we can look back at our solution once more to grasp the logic behind it. We
observed that whatever grammatical string we have will consist of three
portions, the first and last of which can be empty.

\begin{itemize}
\item[]First portion: a string of even length with even number of $b$'s;
crucially a string where every $ab$ or $ba$ we encounter is matched by another
$ab$ or $ba$.
\item[] Second portion: one $ab$ or $ba$, unmatched.
\item[] Third portion: same as the first.
\end{itemize}


Therefore the expression $(ab \cup ba)$ we wrote in (\ref{middle}) corresponds
to the unmatched pair in second portion. Note that some strings may have more
than one analysis into such portions. For instance $abbabaababbb$ can be
broken down in two different ways: (1) $abba\, ba\, ababbb$; (2) $abbabaab\,
ab\, bb$

If you couldn't come up with an answer to the question, please study the
solution very carefully. Or, if you arrived at the same answer as here, try to
find a shorter one.

\end{enumerate}
\end{itemize}
\end{document}



% \item $\setabs{w\in \{a,b\}^*}{w \text{ has at least three } a\text{'s and at least
% two } b\text{'s}}$.



\begin{solution}


\end{solution}
